---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
library(readr)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(xtable)
```

## Loading and preprocessing the data

```{r}
data <- read_csv("activity.csv", col_types = cols(steps = col_integer(), 
                                                date = col_date(), 
                                                interval = col_integer() ))
data_noNA <- filter(data,  !is.na(data$steps)) ## remove NA values
str(data_noNA)
```

## What is mean total number of steps taken per day?

```{r}
steps_day <- data_noNA %>% group_by(date) %>% summarize(total = sum(steps))
head(steps_day,3)
g_his <- ggplot(steps_day, aes(total))+geom_histogram(binwidth = 2000)+
                labs(title = "Histogram. Steps per day.")
g_day <-  ggplot(steps_day, aes(x = date, y = total))+geom_point()+geom_path()+
                labs(title = "Total steps per day")
grid.arrange(g_his, g_day, ncol=2)
summ1 <- summary(steps_day$total)
```
Mean of total number of steps:

`r round(summ1[4],1)`

## What is the average daily activity pattern?

```{r}
## Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, 
## averaged across all days (y-axis)
steps_interval <- data_noNA %>% group_by(interval) %>% summarize(Steps = round(mean(steps),2))
## round(aggregate(steps~interval,data_noNA,mean),2)  ##Other way to do the same
g_day_pattern <- ggplot(steps_interval, aes(x=interval,y=Steps)) + geom_path() + 
        labs(title = "Mean daily steps each 5 min")
g_day_pattern
## Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
steps_interval[which.max(steps_interval$Steps),]
```


## Imputing missing values

```{r}
## Calculate and report the total number of missing values in the dataset
sum(is.na(data$steps))
## I wil use mean to fill missing values. col steps2 contains the mean values
newData <- data %>% group_by(date) %>% mutate(steps2 = steps_interval$Steps)
for(i in 1:length(newData$steps)){
        if(is.na(newData$steps[i])){
                newData$steps[i] <- newData$steps2[i]
        }
}
newData <- select(newData, c(1,2,3))
## New data set
head(newData,2)
## new plots
steps_interval_2 <- newData %>% group_by(interval) %>% summarize(Steps = round(mean(steps),2))
steps_day_2 <- newData %>% group_by(date) %>% summarize(total = sum(steps))
g_his_2 <- ggplot(steps_day_2, aes(total))+geom_histogram(binwidth = 2000)+
                labs(title = "Imputing missing values.")
g_day_2 <-  ggplot(steps_day_2, aes(x = date, y = total))+geom_point()+geom_path()+
                labs(title = "Imputing missing values.")
grid.arrange(g_his,g_day,g_his_2, g_day_2, ncol=2)
summ2 <- summary(steps_day_2$total)
```

Doing a quick summary on the data with (summ2) and without Na values (summ2), we see no changes on 
mean value as spected because of the strategy for imputing Na values.

```{r, results="asis"}
tb <- xtable(data.frame(rbind(summ1,summ2)))
print(tb, type ="html")
```

## Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing 
values for this part.

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given 
date is a weekday or weekend day.

```{r}
newData <- newData %>% mutate(weekday = weekdays(date))
for(i in 1:nrow(newData)){
         newData$weekday[i] <- if(weekdays(newData$date[i]) == "Saturday" | weekdays(newData$date[i]) == "Sunday"){
               "weekend"
        }else{ "weekday" }
}
weeklist <- split(newData, newData$weekday)
```

Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps
taken, averaged across all weekday days or weekend days (y-axis). I think the easiest way to look for any pattern 
change is to superimpose both data patterns:

```{r}
day_interval <- newData %>% group_by(interval,weekday) %>% summarize(Steps = round(mean(steps),2))
g_week <- ggplot(day_interval, aes(x=interval,y=Steps)) 
g_week + facet_grid(weekday~.)+ geom_path(aes(color=weekday))+ labs(title = "Time series plot. 5-min interval")
```